---
title: "Getting started"
---

## About this workshop

This is the **What is a forecast, and is it any good?** workshop for [IDDconf 2026](https://iddconf.com): a short, hands-on selection of forecasting [sessions](timetable) drawn from the fuller MIT-licensed course on [Nowcasting and forecasting infectious disease dynamics](https://nfidd.github.io/nfidd/).
It is designed to help epidemiologists, modellers and researchers visualise, evaluate and combine forecasts of infectious disease surveillance data.

## Set up

Each session in this course uses R code for demonstration. All the content is self-contained within a software package designed for the course.

**You have three options for using this course:**

1.  **Web-only**: View sessions on the website
2.  **Local setup**: Install R, packages, and download course materials for the full interactive experience
3.  **Hybrid**: Install just the packages but use the website for viewing content

::: callout-important
**Installation Issues?** If you're having trouble with any installation steps, **ask for help early**! Don't skip ahead - each step builds on the previous ones. On the day of the course, we have a small number of web clients available as backup if installation issues persist.
:::

### Summary of Installation Steps

If you choose the local setup option, here's what you'll need to do:

1.  Install R and RStudio (use the Visual Editor for best notebook experience)
2.  Install the `iddconf2026` R package
3.  Download workshop materials (if using the hybrid approach you may not need to do this)

**Don't skip any steps** - they all work together to provide the full workshop experience.

::: callout-note
This workshop works entirely from pre-generated forecasts, so you do **not** need to install `cmdstan` or a C++ toolchain.
Everything runs from the `iddconf2026` R package and a few plotting and scoring packages.
:::

### Installing R

-   [R](http://cran.r-project.org) is used as the main programming language.
	You can check which version you have by typing `R.version` in your R session.
	We recommend installing the latest `r R.version$version.string`.
-   [RStudio](http://www.rstudio.com/products/rstudio/download/) is a popular graphic user interface (GUI). 
	Its Visual Editor provides the best experience of going through this course. 
	Please make sure you update RStudio to the latest version.

### Installing additional requirements

Before you get started with the course, you will first need to install the following software.

#### Installation of the `iddconf2026` package

The workshop materials live in the `iddconf2026` package, which contains the data files and helper functions used throughout.
It is installed from GitHub using the `pak` package.

First install `pak` if you do not already have it:

```{r install-pak, eval=FALSE}
install.packages("pak")
```

Then install the `iddconf2026` package and all of its dependencies from GitHub:

```{r install, eval=FALSE}
pak::pak("nfidd/iddconf2026", dependencies = TRUE)
```

Then you can check that the installation completed successfully by loading the package into your **R** session:

```{r load, eval=FALSE}
library("iddconf2026")
```

## Accessing the course

If you want to use the local workflow, you will need a local copy of the course material.

-   Directly download the course material:

    ::: callout-tip
    [[**Download**]{.underline}](https://github.com/nfidd/iddconf2026/archive/refs/heads/main.zip)
    :::

-   Alternatively, if you are familiar with git you can clone the [repo](https://github.com/nfidd/iddconf2026).

-   If you prefer to use the hybrid workflow, you can view each session on the website, and copy-paste the code into your own R script. 
	In that case you don't need to download the material.

    -   Tip: if you hover over each code chunk on the website you can use a "Copy" button at the top right corner.

### Interacting with a local copy of the course material

A benefit of downloading or cloning all the material is that you can interact with the session files directly.

In this course, all content is written using [Quarto](https://quarto.org) notebooks (`.qmd` files).
This means that we can combine text with code and see the output directly. 
The notebooks are then directly reproduced on the course website (for example, this page).

**Recommended approach**: Work with the notebooks using RStudio's visual editor mode. See guidance on this below.

::: callout-tip
## Using RStudio's Visual Editor (Recommended for Notebooks)

1.  **Open a session notebook**: Each session is saved as a `.qmd` file in `nfidd/sessions/`
2.  **Switch to Visual mode**: Look for the "Visual" button in the top-left of the editor pane (next to "Source")
3.  **Execute code**: Use the green "play" button at the top-right corner of each code chunk, or `Ctrl/Cmd + Enter` for line-by-line execution
4.  **Visual mode benefits**: Easier to read formatted text and equations, better experience with code chunks and outputs
:::

**Alternative approaches that also work**:

**Other editors**: Use VS Code or other editors that support Quarto notebooks. The `.qmd` files will work in any Quarto-compatible environment.

::: callout-tip
The Quarto extension for VS Code also supports a visual editor mode.
You can find it in the command palette.
:::

## Day-of-Course Updates

If you're returning to the course after some time or joining a live session, you may want to update your setup to ensure you have the latest content and package versions.

### Quick Update (Recommended)

1.  **Update the `iddconf2026` package** (this is quick and ensures you have the latest functions):

    ```{r update_package, eval=FALSE}
    pak::pak("nfidd/iddconf2026", dependencies = TRUE)
    ```
2.  **Download fresh course materials** if using local files:

    -   Download the latest version: [[**Download**]{.underline}](https://github.com/nfidd/iddconf2026/archive/refs/heads/main.zip)
    -   Or use `git pull` if you cloned the repository
