---
title: "Forecast ensembles"
order: 9
bibliography: ../nfidd.bib
---

# Introduction

As we saw in the [forecasting models](forecasting-models) session, different modelling approaches have different strengths and weaknesses, and we do not usually know in advance which one will produce the best forecast in any given situation.
One way to draw strength from a diversity of approaches is to combine forecasts from different models into a forecast ensemble.

We will build ensembles from the models we met earlier before returning to the question: **is it any good?**, but this time considering what / for whom an ensemble might be good for.

## Slides

- [Introduction to ensembles](slides/introduction-to-ensembles)

## Objectives

By the end of this session you will be able to:

1. **construct** a simple unweighted and weighted ensemble using R;
2. **compare** ensemble scores to their component models and explain why weighting rarely wins;
3. **interpret** the politics of collaborative modelling and assess the costs and benefits of different forecasting models

::: {.callout-note collapse="true"}
# Setup

## Source file

The source file of this session is located at `sessions/forecast-ensembles.qmd`.

## Libraries used

In this session we will use the `iddconf2026` package to load a data set of infection times and forecasts, the `dplyr` and `tidyr` packages for data wrangling, the `ggplot2` library for plotting and the `scoringutils` package for evaluating forecasts.
We will also use `lopensemble` for mixture ensembles (also called linear opinion pool).

```{r}
#| label: libraries
#| message: false

library("iddconf2026")
library("dplyr")
library("tidyr")
library("ggplot2")
library("scoringutils")
library("lopensemble")
```

::::: {.callout-tip}
The best way to interact with the material is via the [Visual Editor](https://docs.posit.co/ide/user/ide/guide/documents/visual-editor.html) of RStudio.
:::::

## Initialisation

We set a random seed for reproducibility.
Setting this ensures that you should get exactly the same results on your computer as we do.

```{r}
set.seed(123)
```
:::

::: {.callout-note .optional collapse="true"}
## Representations of probabilistic forecasts

Probabilistic predictions can be described as coming from a probability distribution.
In general, and when using complex models such as the ones in this course, these distributions cannot be expressed in a simple analytical form.
Instead, we typically use a limited number of samples generated from Monte-Carlo methods to represent the predictive distribution.
However, this is not the only way to characterise a distribution.

A quantile is the value that corresponds to a given quantile level of a distribution.
For example, the median is the 50th quantile: 50% of the distribution lies below it.
If we characterise a predictive distribution by its quantiles, we specify these values at a range of quantile levels, e.g. from 5% to 95% in 5% steps.

Many collaborative forecasting projects and so-called forecasting hubs use quantile-based representations of forecasts, in the hope of characterising both the centre and the tails of the distribution reliably and with less demand on storage than a sample-based representation.
:::

# Individual forecast models

We will use the forecasts from the models we explored in the [forecasting models](forecasting-models) session.
They share the same renewal-with-delays structure but use different models for the evolution of the reproduction number over time:

- A random walk model, the baseline model (`rw_forecasts`)
- A differenced autoregressive model referred to as "More statistical" (`stat_forecasts`)
- A simple model of susceptible depletion referred to as "More mechanistic" (`mech_forecasts`)

The precise details of the models are not critical to what we are doing here.
As in the session on [forecast evaluation](forecast-evaluation), we fitted these models to a range of forecast dates so you don't have to wait for them to fit.

```{r}
#| label: load_forecasts

data(rw_forecasts, stat_forecasts, mech_forecasts)
forecasts <- bind_rows(
  rw_forecasts,
  mutate(stat_forecasts, model = "More statistical"),
  mutate(mech_forecasts, model = "More mechanistic")
)

forecasts
```

::: {.callout-tip .optional collapse="true"}
## How did we generate these forecasts?

We generated these forecasts using the code in `data-raw/generate-example-forecasts.r`.
Some important things to note about them:

- We used a 14 day forecast horizon.
- Each forecast used all the data up to the forecast date.
- We generated 1000 predictive posterior samples for each forecast.
- We started forecasting 3 weeks into the outbreak and then forecast every 7 days until the end of the data (excluding the last 14 days to allow a full forecast).
- We use the same simulated outbreak data as before:

```{r}
gen_time_pmf <- make_gen_time_pmf()
ip_pmf <- make_ip_pmf()
onset_df <- simulate_onsets(
  make_daily_infections(infection_times), ip_pmf
)
head(onset_df)
```
:::

# Forecast ensembles

We now combine multiple forecasts into one.
This is sometimes called *stacking*, and the resulting forecasts are said to come from an *ensemble* of models.

## Quantile-based ensembles

We first consider forecasts based on the individual quantiles of each model.
This treats each model as an uncertain estimate of a single true predictive distribution.
By averaging across models, we aim for a better estimate of that distribution than any single model gives.
If we believe some models are better than others, we can create a weighted version of this average.

### Converting sample-based forecasts to quantile-based forecasts

We will think about forecasts in terms of quantiles, so we first convert our sample-based forecasts to quantile-based ones.
We do this by focusing on the *marginal distribution* at each time point: we treat each time point as independent and calculate quantiles from the sample trajectories at that point.
The `{scoringutils}` package makes this easy.
First we declare the forecasts as `sample` forecasts.

```{r}
#| label: convert-for-scoringutils

sample_forecasts <- forecasts |>
  left_join(onset_df, by = "day") |>
  filter(!is.na(.value)) |>
  as_forecast_sample(
    forecast_unit = c("origin_day", "horizon", "model", "day"),
    observed = "onsets",
    predicted = ".value",
    sample_id = ".draw"
  )
sample_forecasts
```

Then we convert to `quantile` forecasts.

```{r}
#| label: convert-to-quantile

quantile_forecasts <- sample_forecasts |>
  as_forecast_quantile()
quantile_forecasts
```

::: {.callout-tip .optional collapse="true"}
#### What is happening here?

- Internally `scoringutils` calculates the quantiles of the sample-based forecasts.
- It uses a set of default quantile levels, which the user can override.
- It calls base R's `quantile()` function, estimating the value at each quantile level by ordering the samples and taking the value at the appropriate position.
:::

### Simple unweighted ensembles

A good place to start is to take the mean or median across models at each quantile level, and treat the result as the quantiles of the ensemble.
The median is preferred when outlier forecasts are likely, as it is less sensitive to them.
The mean is preferred when we want models that diverge from the middle to still pull the ensemble.

::: {.callout-note .optional collapse="true"}
#### Vincent average

Calculating quantiles of a new distribution as an average of the quantiles of its constituents is called a *Vincent average*, after the biologist Stella Vincent, who described it as early as 1912 when studying the role of whiskers in the behaviour of white rats.
:::

#### Construction

We calculate the mean quantile ensemble by taking the mean of the forecasts at each quantile level.

```{r}
#| label: mean-ensemble

mean_ensemble <- quantile_forecasts |>
  as_tibble() |>
  summarise(
    predicted = mean(predicted),
    observed = unique(observed),
    model = "Mean ensemble",
    .by = c(origin_day, horizon, quantile_level, day)
  )
```

Similarly, the median ensemble takes the median at each quantile level.

```{r}
#| label: median-ensemble

median_ensemble <- quantile_forecasts |>
  as_tibble() |>
  summarise(
    predicted = median(predicted),
    observed = unique(observed),
    model = "Median ensemble",
    .by = c(origin_day, horizon, quantile_level, day)
  )
```

We combine the ensembles with the individual forecasts to make visualisation easier.

```{r}
#| label: combine-ensembles

simple_ensembles <- bind_rows(
  mean_ensemble,
  median_ensemble,
  quantile_forecasts
)
```

#### Visualisation

How do these ensembles differ from the individual models?
We start by plotting a single forecast from each.

```{r}
#| label: plot-single-forecasts

plot_ensembles <- function(data, obs_data) {
  data |>
    pivot_wider(names_from = quantile_level, values_from = predicted) |>
    ggplot(aes(x = day)) +
    geom_ribbon(
      aes(
        ymin = .data[["0.05"]], ymax = .data[["0.95"]], fill = model,
        group = origin_day
      ),
      alpha = 0.2
    ) +
    geom_ribbon(
      aes(
        ymin = .data[["0.25"]], ymax = .data[["0.75"]], fill = model,
        group = origin_day
      ),
      alpha = 0.5
    ) +
    geom_point(
      data = obs_data,
      aes(x = day, y = onsets), color = "black"
    ) +
    scale_color_binned(type = "viridis") +
    facet_wrap(~model) +
    theme(legend.position = "none")
}

plot_single_forecasts <- simple_ensembles |>
  filter(origin_day == 57) |>
  plot_ensembles(onset_df |> filter(day >= 57, day <= 57 + 14))

plot_single_forecasts
```

The log scale gives another perspective.

```{r}
#| label: plot-single-forecasts-log

plot_single_forecasts +
  scale_y_log10()
```

Now a range of forecasts from each model and ensemble.

```{r}
#| label: plot-multiple-forecasts

plot_multiple_forecasts <- simple_ensembles |>
  plot_ensembles(onset_df |> filter(day >= 21)) +
  lims(y = c(0, 400))

plot_multiple_forecasts
```

Both simple ensembles are less variable than the statistical models but more variable than the mechanistic model, and both look more like the statistical models.
The mean ensemble has slightly tighter uncertainty than the median ensemble, and is more variable across forecast dates, especially after the peak.

#### Evaluation

As in the [forecast evaluation session](forecast-evaluation), we score the ensembles with the `{scoringutils}` `score()` function.

```{r}
#| label: score-ensembles

ensemble_scores <- simple_ensembles |>
  as_forecast_quantile(forecast_unit = c("origin_day", "horizon", "model")) |>
  score()
```

::: {.callout-note}
The **weighted interval score (WIS)** [@bracherEvaluatingEpidemicForecasts2021] is a proper scoring rule for quantile forecasts that approximates the **Continuous Ranked Probability Score (CRPS)** by a weighted sum of prediction intervals.
As the number of intervals grows, the WIS converges to the CRPS, combining sharpness with penalties for over- and under-prediction.

We use it here because we are scoring quantiles, not samples, so we cannot use the CRPS directly.
:::

A high-level overview of the scores by model:

```{r}
#| label: score-overview

ensemble_scores |>
  summarise_scores(by = c("model"))
```

::: {.callout-tip}
##### Take 5 minutes

Which model is best overall, and what other breakdowns (by date, by horizon) might change that answer?
:::

::: {.callout-note collapse="true"}
##### Solution

- The mean ensemble is usually the best performing ensemble overall, but the more mechanistic model often remains the single best model.
- Performance can vary a lot across forecast dates and horizons, so a single overall number hides a great deal.
:::

::: {.callout-tip .optional collapse="true"}
## Interactive: WIS → CRPS

```{=html}
<iframe class="nfidd-embed" src="../interactive/forecast-evaluation.html?embed=wis" title="WIS interactive" loading="lazy" style="width:100%;height:760px;border:1px solid #e5e7eb;border-radius:10px;background:#fff"></iframe>
<script>(function(){if(window.__nfiddEmbedInit)return;window.__nfiddEmbedInit=1;window.addEventListener('message',function(ev){var h=ev&&ev.data&&ev.data.nfiddEmbedHeight;if(!h)return;document.querySelectorAll('iframe.nfidd-embed').forEach(function(f){if(f.contentWindow===ev.source)f.style.height=(h+24)+'px';});});})();</script>
```

Add prediction intervals and watch the WIS converge to the CRPS, with the same dispersion / over / under split.
:::

## Weighted ensembles

The simple mean and median treat every model the same.
We could instead weight the models, giving more weight to those that have forecast well.
We will look at two common weighting methods:

- **Inverse WIS weighting**: weight each model by the inverse of its past WIS.
  Simple to understand and implement, but it does not optimise the weights to produce the best forecast; it just hopes that giving more weight to better past performers helps.
- **A linear opinion pool (LOP)**: a weighted *mixture* of the models' distributions, with weights tuned to optimise the CRPS from samples.
  This interprets the models as competing versions of the truth, rather than noisy estimates of one shared distribution.

::: {.callout-tip}
##### Your turn: work in pairs (10--12 minutes)

Run both weighting methods below (each needs the other later, so run them both), then pair up and take one method each to make sense of.

- **Partner A**: follow the **inverse WIS** weighting and be ready to explain how its weights are set.
- **Partner B**: follow the **linear opinion pool** and be ready to explain how its weights are set.

Then compare at the weights plot and score tables that follow: do your two methods agree on which models matter, and does either clearly beat the simple mean ensemble?
:::

#### Inverse WIS weighting

We first need the WIS for each model, which we already have.
We weight by the inverse WIS over the period *before* each forecast, so we only use information available at the time.

```{r}
#| label: calc-wis

model_scores <- quantile_forecasts |>
  score()
cumulative_scores <- list()
## filter for scores up to the origin day of the forecast
for (day_by in unique(model_scores$origin_day)) {
  cumulative_scores[[as.character(day_by)]] <- model_scores |>
    filter(origin_day < day_by) |>
    summarise_scores(by = "model") |>
    mutate(day_by = day_by)
}
weights_per_model <- bind_rows(cumulative_scores) |>
  select(model, day_by, wis) |>
  mutate(inv_wis = 1 / wis) |>
  mutate(
    inv_wis_total_by_date = sum(inv_wis), .by = day_by
  ) |>
  mutate(weight = inv_wis / inv_wis_total_by_date) |> ## normalise
  select(model, origin_day = day_by, weight)

weights_per_model |>
  pivot_wider(names_from = model, values_from = weight)
```

We apply the weights, using weights from two weeks before each forecast date to inform each ensemble.

```{r}
#| label: apply-weights

inverse_wis_ensemble <- quantile_forecasts |>
  as_tibble() |>
  left_join(
    weights_per_model |>
      mutate(origin_day = origin_day + 14),
    by = c("model", "origin_day")
  ) |>
  # assign equal weights if no weights are available
  mutate(weight = ifelse(is.na(weight), 1 / n_distinct(model), weight)) |>
  summarise(
    predicted = sum(predicted * weight),
    observed = unique(observed),
    model = "Inverse WIS ensemble",
    .by = c(origin_day, horizon, quantile_level, day)
  )
```

#### Linear opinion pool

Instead of averaging quantiles, we can treat the models as different possible versions of the truth and build a weighted *mixture* of their distributions.
This is done from samples, with weights tuned to optimise the CRPS.
The procedure is also called a linear opinion pool.
We use the `lopensemble` package.

Fitting the mixture weights optimises for every forecast date, which is slow with 1000 samples.
For demonstration here, we'll thin to 200 samples per forecast: this is about four times faster and gives essentially the same weights.

```{r}
#| label: mixture

## lopensemble expects a "date" column indicating the timing of forecasts
## thin to 200 samples per forecast to speed up weight estimation
lop_forecasts <- sample_forecasts |>
  as_tibble() |>
  filter(sample_id <= 200) |>
  rename(date = day)

lop_by_forecast <- function(
  sample_forecasts,
  forecast_dates,
  group = c("target_end_date"),
  ...
) {
  lapply(forecast_dates, \(x) {
    y <- sample_forecasts |>
      mutate(target_end_date = x) |>
      dplyr::filter(origin_day <= x) |>
      dplyr::filter(origin_day >= x - (3 * 7 + 1)) |>
      dplyr::filter(origin_day == x | date <= x)
    lop <- mixture_from_samples(y) |>
      filter(date > x) |>
      mutate(origin_day = x)
    return(lop)
  })
}

forecast_dates <- quantile_forecasts |>
  as_tibble() |>
  pull(origin_day) |>
  unique()

lop_ensembles_obj <- lop_by_forecast(
  lop_forecasts,
  forecast_dates[-1]
)

lop_weights <- seq_along(lop_ensembles_obj) |>
  lapply(\(i) attr(lop_ensembles_obj[[i]], "weights") |>
    mutate(origin_day = forecast_dates[i + 1])
  ) |>
  bind_rows()

## combine and generate quantiles from the resulting samples
lop_ensembles <- lop_ensembles_obj |>
  bind_rows() |>
  mutate(model = "Linear Opinion Pool") |>
  rename(day = date) |> ## rename column back for later plotting
  as_forecast_sample() |>
  as_forecast_quantile()
```

### Compare the weighted ensembles

We collect the weighted ensembles alongside the simple ensembles.

```{r}
#| label: combine-weighted-ensembles

weighted_ensembles <- bind_rows(
  inverse_wis_ensemble,
  lop_ensembles,
  simple_ensembles
)
```

#### Visualisation

A range of forecasts from each model and ensemble:

```{r}
#| label: plot-multiple-weighted-forecasts

plot_multiple_weighted_forecasts <- weighted_ensembles |>
  plot_ensembles(onset_df |> filter(day >= 21)) +
  lims(y = c(0, 400))
plot_multiple_weighted_forecasts
```

#### Model weights

We can compare the weights each method assigns to each model over time.

```{r}
#| label: plot-model-weights

weights <- rbind(
  weights_per_model |> mutate(method = "Inverse WIS"),
  lop_weights |> mutate(method = "LOP")
)
weights |>
  ggplot(aes(x = origin_day, y = weight, fill = model)) +
  geom_col(position = "stack") +
  facet_grid(~ method) +
  theme(legend.position = "bottom")
```

::: {.callout-tip}
##### Take 2 minutes

Compare notes with the person next to you.

Do the two methods weight the models differently?
How do the weights change over time?
Are you surprised, given what you know about how the models performed?
:::

::: {.callout-note collapse="true"}
##### Solution

- The two methods can disagree markedly, especially early on when there is little data to learn weights from.
- **Inverse WIS** normalises the inverse of each model's past WIS, so it always spreads weight across all three models and never sets a weight to zero.
  It shifts only gradually as scores accumulate.
- The **linear opinion pool** optimises its mixture weights directly against past performance, so it is freer to concentrate weight on whichever model has been fitting best, and its weights can swing more sharply from one forecast date to the next.
- Look at whether a poorly performing model still receives non-trivial weight.
  If it does, that suggests its poor performance is concentrated in certain phases of the outbreak rather than throughout, so it can still add value at other times.
:::

## Quantitative evaluation

Finally we score the weighted ensembles.evaluate accuracy We drop the first two weeks, which the weighted methods cannot forecast as they need training data.

```{r}
#| label: score-weighted-ensembles

weighted_ensemble_scores <- weighted_ensembles |>
  filter(origin_day >= 29) |>
  as_forecast_quantile(forecast_unit = c("origin_day", "horizon", "model")) |>
  score()

weighted_ensemble_scores |>
  summarise_scores(by = c("model"))
```

Remembering the [forecast evaluation session](forecast-evaluation), we should also check performance on the log scale.

```{r}
#| label: score-overview-weighted-log

log_ensemble_scores <- weighted_ensembles |>
  filter(origin_day >= 29) |>
  as_forecast_quantile(forecast_unit = c("origin_day", "horizon", "model")) |>
  transform_forecasts(
    fun = log_shift,
    offset = 1,
    append = FALSE
  ) |>
  score()

log_ensemble_scores |>
  summarise_scores(by = c("model"))
```

The best weighted ensembles only slightly outperform the simple mean, and sometimes do worse.
Even though this is a simple demonstration, with only three models, this replicates one of the most robust findings in forecasting.

::: {.callout-note .optional collapse="true"}
## Fine-tuning weights

There are many methods for weighting among multiple models, broadly along a spectrum of how crude or fine-tuned the weights are.
Some other examples at each extreme are:

- **Filtering (crudest)** Dropping a model altogether from the ensemble is equivalent to assigning it a weight of zero.
- **Quantile regression averaging (fine-tuned)** Weights are chosen directly to give the best score on past data; this can also be optimised separately per forecast horizon.
  This is the most flexible, and at the most risk of overfitting.

In all methods, there is a trade-off in how much to optimise for past performance, versus how much to allow for the possibility that those models recover in future.
This also involves deciding on the period of past performance that's relevant (e.g. all time, or only recent).
:::

## Quantity, quality, and the combination puzzle

Ensembles do tend to beat their component models in predictive accuracy, but a stubborn result is the "forecast combination puzzle".
A simple, equally weighted average is very hard to beat with more sophisticated weighting [@amaralPostprocessingWeightedCombination2025].
This is because weights have to be estimated, and that estimation is itself uncertain, especially with few past forecasts.
A good working prior is that equal weights are hard to beat.

What does seem to make a difference is adding additional models to the ensemble, which adds variance.
Yet after a point this comes with diminishing returns, because this benefit depends on how independent the models really are.
Two models that share assumptions will add less useful information than two different but plausible views [@fox2024].

::: {.callout-note .optional collapse="true"}
## Combining models from different modellers

Throughout the course we have carefully built models from an understanding of the data-generating process.
Combining these might make sense when we have a clear idea of where variation comes from.
However, when we present work to policy, we often create ensembles combine models from different teams, with different and often unstated assumptions about that process.

- Where does "independence" matter in the modelling process?
:::

## What are ensembles good for?

Ensembles are increasingly how modelling reaches policy.
At the same time, there are two different approaches to what an ensemble is good *for*.

**A product:** The benefits of gaining predictive accuracy are the focus of large, open, collaborative projects such as the US Flu Forecast Hub or the [European Respicast Hub](https://respicast.ecdc.europa.eu/), which collect forecasts in a standard format and combine them into a single ensemble.
This typically involves:

- Creating a standardised, open platform for anyone to contribute
- Collating all models' output based on minimal validation checks
- Combining in an ensemble, which becomes the focus of public communication

Typically, there is relatively little opportunity for interaction between modellers, and while each model may be visualised, the resulting ensemble is the focus output.
The implicit value is the *product*: an ensemble forecast that can be scored, visualised, and viewed by anyone.

**A process:** An alternative view is to focus on the collaborative process of creating an ensemble from multiple modellers.
For example, in the UK during COVID-19, the SPI-M-O group of modellers provided consensus equally-averaged forecasts as part of a qualitative narrative presented to policy-makers[@medley2022].

This approach typically involves

- Targeted recruitment among known modellers
- Scheduled opportunities for interaction, including model comparison, identifying differences, and reaching consensus
- An ensemble with shared ownership of the result.

::: {.callout-tip}
##### Take 2 minutes

Consider the process or product approach to collaborative modelling.

- What might each approach imply for the ensemble weighting schemes we've considered so far?
:::

::: {.callout-tip}
##### Group discussion: product and process

Consider the process or product approach to collaborative modelling.
Discuss in pairs, each taking the stance of a forecast producer or a forecast user.

- What are some of the benefits and the costs of each approach, focusing on the process or the product of collaborative modelling?
:::

## What's good enough?

In this workshop, we have emphasised proper scoring rules with the "forecasting paradigm": sharpness subject to calibration.
An ensemble makes the limits of using a single score obvious.

An ensemble might be the most performant, but still worse to use: harder to explain, slower to produce, entirely dependent on the consistency and quality of the component models.
On the other hand, an ensemble might be the most useful: a reasonably reliable product that gives greater opportunity for modeller-policy interaction.
This points towards a wider question that we haven't touched the sides of here: a good forecast may be not only sharp and calibrated, but also trusted, understood, and relevant to the decision at hand.

# Wrap up

- Review what you've learned in this session with the [learning objectives](../reference/learning_objectives)
- Share your [questions and thoughts](../reference/help)

## References

::: {#refs}
:::
