---
title: "Forecast ensembles"
order: 9
bibliography: ../nfidd.bib
---

# Introduction

As we saw in the [forecasting models](forecasting-models) session, different modelling approaches have different strengths and weaknesses, and we do not usually know in advance which one will produce the best forecast in any given situation.
One way to draw strength from a diversity of approaches is to combine forecasts from different models into a forecast ensemble.

In this session, we'll start with forecasts from the models we explored in the [forecasting models](forecasting-models) session and build ensembles of these models.
We will then compare the performance of these ensembles to the individual models and to each other.
Last, we'll return to the question *is it any good?*, considering how an ensemble might be good for different users and use-cases.

## Slides

- [Introduction to ensembles](slides/introduction-to-ensembles)

## Objectives

The aim of this session is to introduce the concept of ensembles of forecasts and to evaluate the performance of ensembles of the models we explored in the [forecasting models](forecasting-models) session.

::: {.callout-note collapse="true"}
# Setup

## Source file

The source file of this session is located at `sessions/forecast-ensembles.qmd`.

## Libraries used

In this session we will use the `iddconf2026` package to load a data set of infection times and forecasts, the `dplyr` and `tidyr` packages for data wrangling, the `ggplot2` library for plotting and the `scoringutils` package for evaluating forecasts.
We will also use `lopensemble` for mixture ensembles (also called linear opinion pool).

```{r libraries, message = FALSE}
library("iddconf2026")
library("dplyr")
library("tidyr")
library("ggplot2")
library("scoringutils")
library("lopensemble")
```

::::: {.callout-tip}
The best way to interact with the material is via the [Visual Editor](https://docs.posit.co/ide/user/ide/guide/documents/visual-editor.html) of RStudio.
:::::

## Initialisation

We set a random seed for reproducibility.
Setting this ensures that you should get exactly the same results on your computer as we do.

```{r}
set.seed(123)
```
:::

::: {.callout-note .optional collapse="true"}
## Representations of probabilistic forecasts

Probabilistic predictions can be described as coming from a probability distribution.
In general, and when using complex models such as the ones in this course, these distributions cannot be expressed in a simple analytical form.
Instead, we typically use a limited number of samples generated from Monte-Carlo methods to represent the predictive distribution.
However, this is not the only way to characterise a distribution.

A quantile is the value that corresponds to a given quantile level of a distribution.
For example, the median is the 50th quantile: 50% of the distribution lies below it.
If we characterise a predictive distribution by its quantiles, we specify these values at a range of quantile levels, e.g. from 5% to 95% in 5% steps.

Many collaborative forecasting projects and so-called forecasting hubs use quantile-based representations of forecasts, in the hope of characterising both the centre and the tails of the distribution reliably and with less demand on storage than a sample-based representation.
:::

# Individual forecast models

We will use the forecasts from the models we explored in the [forecasting models](forecasting-models) session.
They share the same renewal-with-delays structure but use different models for the evolution of the reproduction number over time:

- A random walk model, the baseline model (`rw_forecasts`)
- A differenced autoregressive model referred to as "More statistical" (`stat_forecasts`)
- A simple model of susceptible depletion referred to as "More mechanistic" (`mech_forecasts`)

The precise details of the models are not critical to what we are doing here.
As in the session on [forecast evaluation](forecast-evaluation), we fitted these models to a range of forecast dates so you don't have to wait for them to fit.

```{r load_forecasts}
data(rw_forecasts, stat_forecasts, mech_forecasts)
forecasts <- bind_rows(
  rw_forecasts,
  mutate(stat_forecasts, model = "More statistical"),
  mutate(mech_forecasts, model = "More mechanistic")
)

forecasts
```

::: {.callout-tip .optional collapse="true"}
## How did we generate these forecasts?

We generated these forecasts using the code in `data-raw/generate-example-forecasts.r`.
Some important things to note about them:

- We used a 14 day forecast horizon.
- Each forecast used all the data up to the forecast date.
- We generated 1000 predictive posterior samples for each forecast.
- We started forecasting 3 weeks into the outbreak and then forecast every 7 days until the end of the data (excluding the last 14 days to allow a full forecast).
- We use the same simulated outbreak data as before:

```{r}
gen_time_pmf <- make_gen_time_pmf()
ip_pmf <- make_ip_pmf()
onset_df <- simulate_onsets(
  make_daily_infections(infection_times), ip_pmf
)
head(onset_df)
```
:::

# Forecast ensembles

We will now move to creating forecasts as a combination of multiple forecasts.
This procedure is also sometimes called _stacking_, and the resulting forecasts are said to come from _ensembles_ of forecast models.

## Quantile-based ensembles

We first consider forecasts based on the individual quantiles of each model.
This treats each model as an uncertain estimate of a single true predictive distribution.
By averaging across models, we aim for a better estimate of that distribution than any single model gives.
If we believe some models are better than others, we can create a weighted version of this average.

### Converting sample-based forecasts to quantile-based forecasts

We will think about forecasts in terms of quantiles, so we first convert our sample-based forecasts to quantile-based ones.
We do this by focusing on the *marginal distribution* at each time point: we treat each time point as independent and calculate quantiles from the sample trajectories at that point.
The `{scoringutils}` package makes this easy.
First we declare the forecasts as `sample` forecasts.

```{r convert-for-scoringutils}
sample_forecasts <- forecasts |>
  left_join(onset_df, by = "day") |>
  filter(!is.na(.value)) |>
  as_forecast_sample(
    forecast_unit = c("origin_day", "horizon", "model", "day"),
    observed = "onsets",
    predicted = ".value",
    sample_id = ".draw"
  )
sample_forecasts
```

Then we convert to `quantile` forecasts.

```{r convert-to-quantile}
quantile_forecasts <- sample_forecasts |>
  as_forecast_quantile()
quantile_forecasts
```

::: {.callout-tip .optional collapse="true"}
#### What is happening here?

- Internally `scoringutils` calculates the quantiles of the sample-based forecasts.
- It uses a set of default quantile levels, which the user can override.
- It calls base R's `quantile()` function, estimating the value at each quantile level by ordering the samples and taking the value at the appropriate position.
:::

### Simple unweighted ensembles

A good place to start is to take the mean or median across models at each quantile level, and treat the result as the quantiles of the ensemble.
The median is preferred when outlier forecasts are likely, as it is less sensitive to them.
The mean is preferred when we want models that diverge from the middle to still pull the ensemble.

::: {.callout-note .optional collapse="true"}
#### Vincent average

Calculating quantiles of a new distribution as an average of the quantiles of its constituents is called a *Vincent average*, after the biologist Stella Vincent, who described it in 1912 when studying the role of whiskers in the behaviour of white rats.
:::

#### Construction

We calculate the mean quantile ensemble by taking the mean of the forecasts at each quantile level.

```{r mean-ensemble}
mean_ensemble <- quantile_forecasts |>
  as_tibble() |>
  summarise(
    predicted = mean(predicted),
    observed = unique(observed),
    model = "Ensemble (mean)",
    .by = c(origin_day, horizon, quantile_level, day)
  )
```

Similarly, the median ensemble takes the median at each quantile level.

```{r median-ensemble}
median_ensemble <- quantile_forecasts |>
  as_tibble() |>
  summarise(
    predicted = median(predicted),
    observed = unique(observed),
    model = "Ensemble (median)",
    .by = c(origin_day, horizon, quantile_level, day)
  )
```

We combine the ensembles with the individual forecasts to make visualisation easier.

```{r combine-ensembles}
simple_ensembles <- bind_rows(
  mean_ensemble,
  median_ensemble,
  quantile_forecasts
)
```

#### Visualisation

How do these ensembles differ from the individual models?
We start by plotting a single forecast from each.

```{r plot-single-forecasts}
plot_ensembles <- function(data, obs_data) {
  data |>
    pivot_wider(names_from = quantile_level, values_from = predicted) |>
    ggplot(aes(x = day)) +
    geom_ribbon(
      aes(
        ymin = .data[["0.05"]], ymax = .data[["0.95"]], fill = model,
        group = origin_day
      ),
      alpha = 0.2
    ) +
    geom_ribbon(
      aes(
        ymin = .data[["0.25"]], ymax = .data[["0.75"]], fill = model,
        group = origin_day
      ),
      alpha = 0.5
    ) +
    geom_point(
      data = obs_data,
      aes(x = day, y = onsets), color = "black"
    ) +
    scale_color_binned(type = "viridis") +
    facet_wrap(~model) +
    theme(legend.position = "none")
}

plot_single_forecasts <- simple_ensembles |>
  filter(origin_day == 57) |>
  plot_ensembles(onset_df |> filter(day >= 57, day <= 57 + 14))

plot_single_forecasts
```

The log scale gives another perspective.

```{r plot-single-forecasts-log}
plot_single_forecasts +
  scale_y_log10()
```

Now the full range of forecasts from each model and ensemble.

```{r plot-multiple-forecasts}
plot_multiple_forecasts <- simple_ensembles |>
  plot_ensembles(onset_df |> filter(day >= 21)) +
  lims(y = c(0, 400))

plot_multiple_forecasts
```

::: {.callout-tip}
## Take 2 minutes
How do these ensembles compare to the individual models?
How do they differ from each other, and are there any differences across forecast dates?
:::

::: {.callout-note collapse="true"}
## Solution
How do these ensembles compare to the individual models?

- Both simple ensembles appear to be less variable than the statistical models but more variable than the mechanistic model.
- Both ensembles are more like the statistical models than the mechanistic model.

How do they differ from each other, and are there any differences across forecast dates?

- The mean ensemble has slightly tighter uncertainty than the median ensemble.
- The mean ensemble is more variable across forecast dates, especially after the peak of the outbreak.
:::

#### Evaluation

As in the [forecast evaluation session](forecast-evaluation), we score the ensembles with the `{scoringutils}` `score()` function.

```{r score-ensembles}
ensemble_scores <- simple_ensembles |>
  as_forecast_quantile(forecast_unit = c("origin_day", "horizon", "model")) |>
  score()
```

::: {.callout-note}
The **weighted interval score (WIS)** [@bracherEvaluatingEpidemicForecasts2021] is a proper scoring rule for quantile forecasts that approximates the **Continuous Ranked Probability Score (CRPS)** by a weighted sum of prediction intervals.
As the number of intervals increases, the WIS converges to the CRPS, combining sharpness with penalties for over- and under-prediction.

We use it here because we are scoring quantiles, not samples, so we cannot use the CRPS directly as we did before.
:::

::: {.callout-tip .optional collapse="true"}
## Interactive: WIS → CRPS

```{=html}
<iframe class="nfidd-embed" src="../interactive/forecast-evaluation.html?embed=wis" title="WIS interactive" loading="lazy" style="width:100%;height:760px;border:1px solid #e5e7eb;border-radius:10px;background:#fff"></iframe>
<script>(function(){if(window.__nfiddEmbedInit)return;window.__nfiddEmbedInit=1;window.addEventListener('message',function(ev){var h=ev&&ev.data&&ev.data.nfiddEmbedHeight;if(!h)return;document.querySelectorAll('iframe.nfidd-embed').forEach(function(f){if(f.contentWindow===ev.source)f.style.height=(h+24)+'px';});});})();</script>
```

Add prediction intervals and watch the WIS converge to the CRPS, with the same dispersion / over / under split.
:::

::: {.callout-tip .optional collapse="true"}
## Interactive: interval coverage

```{=html}
<iframe class="nfidd-embed" src="../interactive/forecast-evaluation.html?embed=coverage" title="interval coverage interactive" loading="lazy" style="width:100%;height:720px;border:1px solid #e5e7eb;border-radius:10px;background:#fff"></iframe>
<script>(function(){if(window.__nfiddEmbedInit)return;window.__nfiddEmbedInit=1;window.addEventListener('message',function(ev){var h=ev&&ev.data&&ev.data.nfiddEmbedHeight;if(!h)return;document.querySelectorAll('iframe.nfidd-embed').forEach(function(f){if(f.contentWindow===ev.source)f.style.height=(h+24)+'px';});});})();</script>
```

A calibration check for quantile forecasts: does a 90% prediction interval actually contain the truth 90% of the time?
:::

We start with a high-level overview of the scores by model:

```{r score-overview}
ensemble_scores |>
  summarise_scores(by = c("model"))
```

::: {.callout-tip}
## Take 2 minutes

What do you think the scores are telling you? Which model do you think is best? What other scoring breakdowns might you want to look at?
:::

::: {.callout-note collapse="true"}
## Solution

What do you think the scores are telling you? Which model do you think is best?

- The mean ensemble appears to be the best performing ensemble model overall.
- However, the more mechanistic model remains the single best model among any.

What other scoring breakdowns might you want to look at?

- There might be variation between the performance of different ensemble methods over different forecast dates, or the length of forecast horizon.
:::

## Weighted ensembles

The simple mean and median treat every model the same.
We could instead weight the models, for example giving more weight to those that have forecast well.
As the crudest option, **filtering** simply drops a poorly performing model, which is equivalent to giving it a weight of zero.

Let's look at two common weighting methods:

- **Inverse WIS weighting**: 
  - Use: weight each model by the inverse of its past WIS. 
    Simple to understand and implement; relies on the hope that giving more weight to better past performers yields a better ensemble.
  - Interpretation: each componennt model provides a different uncertain estimates of a true distribution of a given shape.
- **A linear opinion pool (LOP)**: 
  - Use: weights are tuned to optimise the CRPS from samples, creating a weighted mixture of the models' distributions.
  - Interpretation: each component model is a competing version of the truth among multiple possible versions, with weights assigned to each representing the probability of each one being the true one.

::: {.callout-tip}
## Take 5 minutes: work in pairs

Pair up and take one method each to make sense of. 
Note that if you are runnning code interactively, you will still need to run each code chunk.

- **Partner A**: follow the **inverse WIS** weighting and be ready to explain how its weights are set.
- **Partner B**: follow the **linear opinion pool** and be ready to explain how its weights are set.

Compare the weights plot and score tables that follow: do the two weighting methods agree, and does either clearly beat the simple ensembles?
:::

### Inverse WIS weighting

We first need the WIS for each model, which we already have.
We weight by the inverse WIS over the period before each forecast, so we only use information available at the time.

```{r calc-wis}
model_scores <- quantile_forecasts |>
  score()
cumulative_scores <- list()
## filter for scores up to the origin day of the forecast
for (day_by in unique(model_scores$origin_day)) {
  cumulative_scores[[as.character(day_by)]] <- model_scores |>
    filter(origin_day < day_by) |>
    summarise_scores(by = "model") |>
    mutate(day_by = day_by)
}
weights_per_model <- bind_rows(cumulative_scores) |>
  select(model, day_by, wis) |>
  mutate(inv_wis = 1 / wis) |>
  mutate(
    inv_wis_total_by_date = sum(inv_wis), .by = day_by
  ) |>
  mutate(weight = inv_wis / inv_wis_total_by_date) |> ## normalise
  select(model, origin_day = day_by, weight)

weights_per_model |>
  pivot_wider(names_from = model, values_from = weight)
```

We apply the weights, using weights from two weeks before each forecast date to inform each ensemble.

```{r apply-weights}
inverse_wis_ensemble <- quantile_forecasts |>
  as_tibble() |>
  left_join(
    weights_per_model |>
      mutate(origin_day = origin_day + 14),
    by = c("model", "origin_day")
  ) |>
  # assign equal weights if no weights are available
  mutate(weight = ifelse(is.na(weight), 1 / n_distinct(model), weight)) |>
  summarise(
    predicted = sum(predicted * weight),
    observed = unique(observed),
    model = "Ensemble (Inverse WIS)",
    .by = c(origin_day, horizon, quantile_level, day)
  )
```

### Linear opinion pool

We can build a weighted mixture of models' distributions, with sample weights tuned to optimise the CRPS, using the `lopensemble` package.

Fitting the mixture weights optimises for every forecast date, which is slow with 1000 samples.
For demonstration here, we'll thin to 200 samples per forecast: this is about four times faster and gives essentially the same weights.

```{r mixture}
## lopensemble expects a "date" column indicating the timing of forecasts
## thin to 200 samples per forecast to speed up weight estimation
lop_forecasts <- sample_forecasts |>
  as_tibble() |>
  filter(sample_id <= 200) |>
  rename(date = day)

lop_by_forecast <- function(
  sample_forecasts,
  forecast_dates,
  group = c("target_end_date"),
  ...
) {
  lapply(forecast_dates, \(x) {
    y <- sample_forecasts |>
      mutate(target_end_date = x) |>
      dplyr::filter(origin_day <= x) |>
      dplyr::filter(origin_day >= x - (3 * 7 + 1)) |>
      dplyr::filter(origin_day == x | date <= x)
    lop <- mixture_from_samples(y) |>
      filter(date > x) |>
      mutate(origin_day = x)
    return(lop)
  })
}

forecast_dates <- quantile_forecasts |>
  as_tibble() |>
  pull(origin_day) |>
  unique()

lop_ensembles_obj <- lop_by_forecast(
  lop_forecasts,
  forecast_dates[-1]
)

lop_weights <- seq_along(lop_ensembles_obj) |>
  lapply(\(i) attr(lop_ensembles_obj[[i]], "weights") |>
    mutate(origin_day = forecast_dates[i + 1])
  ) |>
  bind_rows()

## combine and generate quantiles from the resulting samples
lop_ensembles <- lop_ensembles_obj |>
  bind_rows() |>
  mutate(model = "Ensemble (Linear Opinion Pool)") |>
  rename(day = date) |> ## rename column back for later plotting
  as_forecast_sample() |>
  as_forecast_quantile()
```

::: {.callout-tip .optional collapse="true"}
## Interactive: opinion pool vs quantile averaging

```{=html}
<iframe class="nfidd-embed" src="../interactive/forecast-ensembles.html?embed=headtohead" title="opinion pool vs quantile averaging interactive" loading="lazy" style="width:100%;height:900px;border:1px solid #e5e7eb;border-radius:10px;background:#fff"></iframe>
<script>(function(){if(window.__nfiddEmbedInit)return;window.__nfiddEmbedInit=1;window.addEventListener('message',function(ev){var h=ev&&ev.data&&ev.data.nfiddEmbedHeight;if(!h)return;document.querySelectorAll('iframe.nfidd-embed').forEach(function(f){if(f.contentWindow===ev.source)f.style.height=(h+24)+'px';});});})();</script>
```

Average probabilities (opinion pool) vs quantiles (Vincent) — and watch them diverge.
:::

### Compare the different ensembles

We collect the weighted ensembles alongside the simple ensembles.

```{r combine-weighted-ensembles}
weighted_ensembles <- bind_rows(
  inverse_wis_ensemble,
  lop_ensembles,
  simple_ensembles
)
```

#### Visualisation

A range of forecasts from each model and ensemble:

```{r plot-multiple-weighted-forecasts}
plot_multiple_weighted_forecasts <- weighted_ensembles |>
  plot_ensembles(onset_df |> filter(day >= 21)) +
  lims(y = c(0, 400))
plot_multiple_weighted_forecasts
```

#### Model weights

We can compare the weights each method assigns to each model over time.

```{r plot-model-weights}
weights <- rbind(
  weights_per_model |> mutate(method = "Inverse WIS"),
  lop_weights |> mutate(method = "LOP")
)
weights |>
  ggplot(aes(x = origin_day, y = weight, fill = model)) +
  geom_col(position = "stack") +
  facet_grid(~ method) +
  theme(legend.position = "bottom")
```

::: {.callout-tip}
#### Take 2 minutes
Compare notes with the person next to you.

Are the weights assigned to models different between the three methods?
How do the weights change over time?
Are you surprised by the results given what you know about the models performance?
:::

::: {.callout-note collapse="true"} 
#### Solution
Are the weights assigned to models different between the methods?

- There are major differences especially early on, where the LOP ensemble prefers the random walk model.
- The inverse WIS model transitions from fairly even weights early on to giving most weight to the mechanistic model, however it does so in a more balanced manner than the optimised ensemble, giving substantial weight to all three models.

How do the weights change over time?

- Early on the more statistical models have higher weights in the respective ensemble methods.
- Gradually the mechanistic model gains weight in both models and by the end of the forecast horizon it represents the entire ensemble.

Are you surprised by the results given what you know about the models performance?

- As the random walk model is performing poorly, you would expect it to have low weights but actually it often doesn't.
  This implies that its poor performance is restricted to certain parts of the outbreak.
- The mechanistic model performs really well overall and dominates the final optimised ensemble.
:::

## Evaluation

Finally we evaluate the accuracy of the weighted ensembles.
We drop the first two weeks, which the weighted methods cannot forecast as they need training data.

```{r score-weighted-ensembles}
weighted_ensemble_scores <- weighted_ensembles |>
  filter(origin_day >= 29) |>
  as_forecast_quantile(forecast_unit = c("origin_day", "horizon", "model")) |>
  score()

weighted_ensemble_scores |>
  summarise_scores(by = c("model"))
```

Remembering the [forecast evaluation session](forecast-evaluation), we should also check performance on the log scale.

```{r score-overview-weighted-log}
log_ensemble_scores <- weighted_ensembles |>
  filter(origin_day >= 29) |>
  as_forecast_quantile(forecast_unit = c("origin_day", "horizon", "model")) |>
  transform_forecasts(
    fun = log_shift,
    offset = 1,
    append = FALSE
  ) |>
  score()

log_ensemble_scores |>
  summarise_scores(by = c("model"))
```

The best weighted ensembles only slightly outperform the simple mean, and sometimes do worse.

::: {.callout-note .optional collapse="true"}
## Fine-tuning weights

Any weighting method you use requires a trade-off: how much to optimise for past performance, versus how much to allow that a currently poor model may recover. For example, some choices might be:

- **How many past forecasts do you weight over?** More history stabilises the weights but is slower to adapt; less history adapts quickly but is noisier.
- **Which period counts as relevant?** All time, or only recent weeks — this trades off responsiveness against stability.
- **How flexible should the weights be?** Whether to use weights per-model, per-horizon, or per-quantile; each added degree of freedom can improve the fit on past data but risks overfitting the future.
:::

::: {.callout-note .optional collapse="true"}
## Quantile Regression Averaging

**Quantile regression averaging (QRA)** treats ensemble weighting as a regression problem. 
QRA fits the model weights directly against past observed outcomes to optimise the score. 
Because it is a regression, it can even fit a separate set of weights for each forecast horizon (or even each quantile), letting the ensemble trust different models at short versus long horizons.
Where the linear opinion pool mixes the models' full distributions, QRA works directly on their quantiles. 

It is available in the `qrensemble` package. The code below fits a QRA ensemble for each forecast date, using up to 3 weeks of previous forecasts, but is left unevaluated here for self-study.

```{r qra, eval=FALSE}

qra_by_forecast <- function(
  quantile_forecasts,
  forecast_dates,
  group = c("target_end_date"),
  ...
) {
  lapply(forecast_dates, \(x) {
    quantile_forecasts |>
      mutate(target_end_date = x) |>
      dplyr::filter(origin_day <= x) |>
      dplyr::filter(origin_day >= x - (3 * 7 + 1)) |>
      dplyr::filter(origin_day == x | day <= x) |>
      qrensemble::qra(
        group = group,
        target = c(origin_day = x),
        ...
      )
  })
}

## a single optimised ensemble, shared across all horizons
qra_ensembles_obj <- qra_by_forecast(
  quantile_forecasts,
  forecast_dates[-1],
  group = c("target_end_date")
)

## or a separate optimised ensemble per forecast horizon
qra_by_horizon_obj <- qra_by_forecast(
  quantile_forecasts,
  forecast_dates[-c(1:2)],
  group = c("horizon", "target_end_date")
)
```
:::

::: {.callout-note .optional collapse="true"}
## The forecast combination puzzle

Ensembles do tend to beat their component models in predictive accuracy. 
But it is very hard to beat a simple, equally weighted average, even with more sophisticated weighting procedures [@amaralPostprocessingWeightedCombination2025].
This is known as the "forecast combination puzzle". 
It's because weights have to be estimated, and that estimation is itself uncertain, especially with few past forecasts.
A good working prior is that equal weights are hard to beat.

:::

## What are ensembles good for?

An ensemble might be the most performant, but still harder to use: harder to explain, slower to produce, and sensitive to the consistency of its component models (though, as in machine learning, an ensemble of individually weak models can still perform well together).
On the other hand, an ensemble might be the most useful: a reasonably reliable product that allows decision-makers to focus on a single useable result.
This points towards a wider question that we haven't touched the sides of here: a good forecast should be not only sharp and calibrated, but also trusted, understood, and relevant to the decision at hand. 

This might suggest two different views of what an ensemble might be good for.

**A product:** The benefits of gaining predictive accuracy are the focus of large, open, collaborative projects such as the US Flu Forecast Hub or the [European Respicast Hub](https://respicast.ecdc.europa.eu/).
These projects collate forecasts in a standard format and combine them into a single ensemble.
Typically, while each model may be visualised, the resulting ensemble output is the focus [@lutzApplyingInfectiousDisease2019b].
This typically involves:

- Creating a standardised, open platform for anyone to contribute
- Collating all models' output based on minimal validation checks
- Combining in an ensemble, which becomes the focus of public communication

**A process:** An alternative view is to focus on the collaborative process of creating an ensemble from multiple modellers.
For example, in the UK during COVID-19, the SPI-M-O group of modellers provided consensus equally-averaged forecasts as part of a qualitative narrative presented to policy-makers [@medleyConsensusEvidenceRole2022a].
This approach typically involves:

- Selective recruitment among known modellers
- Structured opportunities for interaction, including model comparison, identifying differences, and reaching consensus
- Shared ownership of any consensus output

# Going further

- How might the "product" or "process" approaches relate to the ensemble weighting schemes we considered above?
- What are some of the benefits of a product-focused approach compared to a process-focused approach, and vice-versa?
- What are some of the drawbacks and costs of each approach?
  How could these be mitigated?
- Are there any features of the outbreak setting that might make one approach more appropriate that the other?

## Methods in the real world

- @howertonContextdependentRepresentationBetweenmodel2023 suggests that the choice of an ensemble method should be informed by an assumption about how to represent uncertainty between models: whether differences between component models is "noisy" variation around a single underlying distribution, or represents structural uncertainty about the system.
- @sherrattPredictivePerformanceMultimodel2023 investigates the performance of different ensembles in the European COVID-19 Forecast Hub.
- @amaralPostprocessingWeightedCombination2025 discusses the challenges in improving on the predictive performance of simpler approaches using weighted ensembles.
- @lutzApplyingInfectiousDisease2019b describes the public health use of ensemble forecasts of influenza in the Unites States.
- @medleyConsensusEvidenceRole2022a discusses the process of convening a group of independent modelling teams to provide a consensus for UK COVID-19 public health policy.

# Wrap up

- Review what you've learned in this session with the [learning objectives](../reference/learning_objectives)
- Share your [questions and thoughts](../reference/help)

## References

::: {#refs}
:::
