---
title: "Forecasting concepts and visualisation"
order: 7
bibliography: ../nfidd.bib
---

# Introduction

Epidemiological forecasts are probabilistic statements about what might happen to population disease burden in the future.
In this session we will look at some simple forecasts made using a commonly used infectious disease model, based on the renewal equation.
We will see how we can visualise such forecasts, and visually compare them to what really happened.

## Slides

- [Introduction to forecasting](slides/introduction-to-forecasting)

## Objectives

The aim of this session is to introduce the concept of forecasting and forecast visualisation using a simple model.

::: {.callout-note collapse="true"}

# Setup

## Source file

The source file of this session is located at `sessions/forecasting-concepts.qmd`.

## Libraries used

In this session we will use the `nfidd` package to load a data set of infection times and corresponding forecasts, the `dplyr` package for data wrangling and the `ggplot2` library for plotting.

```{r libraries, message = FALSE}
library("nfidd")
library("dplyr")
library("ggplot2")
```

::: {.callout-tip}
The best way to interact with the material is via the [Visual Editor](https://docs.posit.co/ide/user/ide/guide/documents/visual-editor.html) of RStudio.
:::

## Initialisation

We set a random seed for reproducibility.
Setting this ensures that you should get exactly the same results on your computer as we do.

```{r}
set.seed(123)
```

:::

# What is forecasting?

Forecasting is the process of making predictions about the future based on past and present data.
In the context of infectious disease epidemiology, forecasting is usually the process of predicting the future course of some metric of infectious disease incidence or prevalence based on past and present data.
Here we focus on forecasting observed data (the number of individuals with new symptom onset) but forecasts can also be made for other quantities of interest such as the number of infections, the reproduction number, or the number of deaths.

The **forecasting paradigm** is to **maximise *sharpness* subject to *calibration*** [@gneiting2007].
Forecasts should aim to have narrow uncertainty (sharpness) while remaining correct (calibration).
Note that sharpness is a property of the forecasts themselves, whereas calibration and other measures of forecast quality (such as bias and accuracy) emerge from comparing forecasts to observed data.
We will return to the forecasting paradigm and how to evaluate forecasts against it in the [forecast evaluation session](forecast-evaluation).

# Forecasts from a renewal equation model

For this workshop we work with forecasts that we made earlier and provide in the `nfidd` package, rather than fitting a model ourselves.
The forecasts come from a renewal equation model, which assumes that the number of infectious people determines the number of future infectious people via the reproduction number $R_t$.
You do not need to understand the model in detail to visualise and evaluate its forecasts, but the optional boxes below describe it for those who are interested.

::: {.callout-note}
# The reproduction number $R_t$

The reproduction number $R_t$ (sometimes called the _effective_ reproduction number) describes the average number of secondary infections caused by a single infectious individual.
It can vary in time and space as a function of differences in population level susceptibility, changes in behaviour, policy, seasonality and so on.
The reproduction number $R_t$ is therefore a more general concept than the _basic_ reproduction number $R_0$, which is the average number of secondary infections caused by a single infectious individual in a completely susceptible population.
:::

::: {.callout-tip collapse="true"}
# The discrete renewal equation (optional)

The _renewal equation_ is a general model of infectious disease transmission which includes the SIR model as a special case.
In its basic form it makes no assumption about the specific processes that cause $R_t$ to have a certain value or to change over time.
Instead it relates the number of infected people in the population, the current value of the reproduction number, and a delay distribution that represents the timings of when individuals infect others relative to when they themselves became infected, the so-called generation time.
Mathematically, it can be written as

$$
I_t = R_t \sum_{i=1}^{g_\mathrm{max}} I_{t-i} g_i
$$

Here, $I_t$ is the number of infected individuals on day $t$, $R_t$ is the current value of the reproduction number and $g_i$ is the probability of a secondary infection occurring $i$ days after the infector became infected themselves, with a maximum $g_\mathrm{max}$.
Expected symptom onsets follow from convolving infections with a delay distribution from infection to onset, and the observed onsets are modelled as Poisson draws around these expected onsets.
:::

::: {.callout-note collapse="true"}
# The geometric random walk model (optional)

The forecasts we look at are based on $R_t$ doing a geometric random walk in time.
We might expect the reproduction number to change smoothly over time (except in situations of drastic change such as a very effective intervention) and to be similar at adjacent time points.
We model this by assuming that the logarithm of the reproduction number does a random walk, which keeps the reproduction number positive and makes changes multiplicative rather than additive.
We can write this model as

$$
\sigma \sim HalfNormal(0, 0.05) \\
$$
$$
\log(R_0) \sim \mathcal{Lognormal}(-0.1, 0.5)
$$
$$
\log(R_t) \sim \mathcal{N}(\log(R_{t-1}), \sigma)
$$

The step size of this geometric random walk, $\sigma$, is estimated from data, and is then used to project $R_t$ into the future.
For more information on this approach to forecasting see, for example, @funk2018real.

## Simulating a geometric random walk

You can have a look at an R function for performing the geometric random walk:

```{r geometric-random-walk}
geometric_random_walk
```

We can use this function to simulate a random walk:

```{r simulate-geometric-walk}
R <- geometric_random_walk(init = 1, noise = rnorm(100), std = 0.1)
data <- tibble(t = seq_along(R), R = exp(R))

ggplot(data, aes(x = t, y = R)) +
  geom_line() +
  labs(title = "Simulated data from a random walk model",
       x = "Time",
       y = "R")
```

You can repeat this multiple times, either with the same parameters or changing some, to get a feeling for what this does.
:::

We will now look at some data from an infectious disease outbreak and forecasts from a renewal equation model.
First, we simulate the observed data using functions from the `nfidd` package.

```{r, load-simulated-onset}
gen_time_pmf <- make_gen_time_pmf()
ip_pmf <- make_ip_pmf()
onset_df <- simulate_onsets(
  make_daily_infections(infection_times), ip_pmf
)
tail(onset_df)
```

This uses a data set of infection times that is included in the R package, and a function to turn these into daily number of observed symptom onsets.
Do not worry too much about the details of this.
The important thing is that we now have a data set `onset_df` of symptom onsets.

::: {.callout-tip collapse="true"}
# How does this work?

If you want to see how this works look at the function code by typing `simulate_onsets` in the console.
:::

# Visualising the forecast

We can now visualise a forecast made from the renewal equation model.
This forecast is provided in the `nfidd` package, which we loaded earlier.
We will first extract the forecast and then plot the forecasted number of symptom onsets alongside the observed number of symptom onsets before the forecast was made.

```{r extract-forecast}
data(rw_forecasts)
horizon <- 14
forecast_day <- 71
forecast <- rw_forecasts |>
  ungroup() |>
  filter(origin_day == forecast_day)

previous_onsets <- onset_df |>
  filter(day <= forecast_day)
```

::: {.callout-tip collapse="true"}
## How did we generate these forecasts?
Some important things to note about these forecasts:

  - We used a 14 day forecast horizon.
  - Each forecast used all the data up to the forecast date.
  - We generated 1000 predictive posterior samples for each forecast.
  - We started forecasting 3 weeks into the outbreak and then forecast every 7 days until the end of the data (excluding the last 14 days to allow a full forecast).
  - We generated these forecasts by fitting the renewal equation model offline; the code is in `data-raw/generate-example-forecasts.r`.
:::

```{r plot_forecast}
forecast |>
  filter(.draw %in% sample(.draw, 100)) |>
  ggplot(aes(x = day)) +
  geom_line(alpha = 0.1, aes(y = .value, group = .draw)) +
  geom_point(
    data = previous_onsets, aes(x = day, y = onsets), color = "black"
  ) +
  labs(
    title = "Symptom onsets",
    subtitle = "Forecast (trajectories) and observed (points)"
  )
```

In this plot, the dots show the data and the lines are forecast trajectories that the model deems plausible and consistent with the data so far.

::: {.callout-tip}
## Take 5 minutes
What do you think of this forecast?
Does it match your intuition of how this outbreak will continue?
Is there another way you could visualise the forecast that might be more informative?

:::

::: {.callout-note collapse="true"}
## Solution
- The forecast mainly predicts things to continue growing as they have been. However, some of the trajectories are going down, indicating that with some probabilities the outbreak might end.
- Based purely on the data and without knowing anything else about the disease and setting, it is hard to come up with an alternative. The model seems sensible given the available data. In particular, uncertainty increases with increasing distance to the data, which is a sign of a good forecasting model.
- Instead of visualising plausible trajectories we could visualise a cone with a central forecast and uncertainty around it. We will look at this in the next session as an alternative.
:::

If we want to know how well a model is doing at forecasting, we can later compare it to the data of what really happened.
If we do this, we get the following plot.

```{r plot_future_forecast}
future_onsets <- onset_df |>
  filter(day <= forecast_day + horizon)

forecast |>
  filter(.draw %in% sample(.draw, 100)) |>
  ggplot(aes(x = day)) +
  geom_line(alpha = 0.1, aes(y = .value, group = .draw)) +
  geom_point(data = future_onsets, aes(x = day, y = onsets), color = "black") +
  labs(
    title = "Symptom onsets",
    subtitle = "Forecast (trajectories) and observed (points)"
  )
```

::: {.callout-tip}
## Take 5 minutes
What do you think now of this forecast?
Did the model do a good job?
Again, is there another way you could visualise the forecast that might be more informative?
:::

::: {.callout-note collapse="true"}
## Solution
- On the face of it the forecast looks very poor with some very high predictions compared to the data.
- Based on this visualisation it is hard to tell if the model is doing a good job but it seems like it is not.
- As outbreaks are generally considered to be exponential processes it might be more informative to plot the forecast on the log scale.

```{r plot_forecast_log}
forecast |>
  filter(.draw %in% sample(.draw, 100)) |>
  ggplot(aes(x = day)) +
  geom_line(alpha = 0.1, aes(y = .value, group = .draw)) +
  geom_point(data = future_onsets, aes(x = day, y = onsets), color = "black") +
  scale_y_log10() +
  labs(title = "Symptom onsets, log scale", subtitle = "Forecast and observed")
```

This should be a lot more informative.
We see that for longer forecast horizons the model is not doing a great job of capturing the reduction in symptom onsets.
However, we can now see that the model seems to be producing very reasonable forecasts for the first week or so of the forecast.
This is a common pattern in forecasting where a model is good at capturing the short term dynamics but struggles with the longer term dynamics.
:::

We managed to learn quite a lot about our model's forecasting limitations just looking at a single forecast using visualisations.
However, what if we wanted to quantify how well the model is doing?
In order to do that we need to look at multiple forecasts from the model.

# Visualising multiple forecasts from a single model

As for a single forecast, our first step is to visualise the forecasts as this can give us a good idea of how well the model is doing without having to calculate any metrics.

```{r plot-all-forecasts}
rw_forecasts |>
  filter(.draw %in% sample(.draw, 100)) |>
  ggplot(aes(x = day)) +
  geom_line(
    aes(y = .value, group = interaction(.draw, origin_day), col = origin_day),
    alpha = 0.1
  ) +
  geom_point(
    data = onset_df |>
      filter(day >= 21),
    aes(x = day, y = onsets), color = "black") +
  scale_color_binned(type = "viridis") +
  labs(title = "Weekly forecasts of symptom onsets over an outbreak",
       col = "Forecast start day")
```

As for the single forecast it may be helpful to also plot the forecast on the log scale.

```{r plot-all-forecasts-log}
rw_forecasts |>
  filter(.draw %in% sample(.draw, 100)) |>
  ggplot(aes(x = day)) +
  geom_line(
    aes(y = .value, group = interaction(.draw, origin_day), col = origin_day),
    alpha = 0.1
  ) +
  geom_point(data = onset_df, aes(x = day, y = onsets), color = "black") +
  scale_y_log10() +
  scale_color_binned(type = "viridis") +
  labs(title = "Weekly symptom onset forecasts: log scale",
       col = "Forecast start day")
```

::: {.callout-tip}
## Take 5 minutes
What do you think of these forecasts?
Are they any good?
How well do they capture changes in trend?
Does the uncertainty seem reasonable?
Do they seem to under or over predict consistently?
Would you visualise the forecast in a different way?
:::

::: {.callout-note collapse="true"}
## Solution
What do you think of these forecasts?

  - We think these forecasts are a reasonable place to start but there is definitely room for improvement.

Are they any good?

  - They seem to do a reasonable job of capturing the short term dynamics but struggle with the longer term dynamics.

How well do they capture changes in trend?

  - There is little evidence of the model capturing the reduction in onsets before it begins to show in the data.

Does the uncertainty seem reasonable?

 - On the natural scale it looks like the model often over predicts. Things seem more balanced on the log scale but the model still seems to be overly uncertain.

Do they seem to under or over predict consistently?

  - It looks like the model is consistently over predicting on the natural scale but this is less clear on the log scale.
:::

# Going further

## Challenge

- What other ways could we visualise the forecasts, especially if we had forecasts from multiple models and locations, or both?
- Based on these visualisations how could we improve the model?

## Methods in practice

- @heldProbabilisticForecastingInfectious2017 makes a compelling argument for the use of probabilistic forecasts and evaluates spatial forecasts based on routine surveillance data.
- @hadleyVisualPreferencesCommunicating2025 describes the visualisation of modelling results for national COVID-19 policy and decision makers, with recommendations on what aspects of visuals, graphs, and plots policymakers found to be most helpful in COVID-19 response work.

::: {.callout-note collapse="true"}
## Using this model as a baseline

The random walk renewal model used here is deliberately simple, which makes it a natural *baseline*: a reference forecast that more sophisticated models should be expected to beat before we trust them.
Comparing models against a baseline is central to forecast evaluation, and the choice of baseline itself can substantially change which models appear skilful [@stapperMindBaseline2025].
Baselines are useful because they:

- give an interpretable reference point for whether a model adds value beyond a naive extrapolation of recent trends;
- are cheap to fit and robust, so they almost always produce a forecast to compare against;
- help reveal when added model complexity is not justified by better performance.

Which of these, if any, does the random walk renewal model used here satisfy?
:::

::: {.callout-note collapse="true"}
## Solution
- Interpretable reference point: yes, it's just a naive extrapolation of the recent trend in $R_t$.
- Cheap to fit and robust: not really. It's a full Bayesian latent-state model that takes appreciably longer to fit than a naive baseline, and it needs careful sampler tuning to sample cleanly.
- Revealing unnecessary complexity: yes, if competing models are compared against it fairly.

Stapper & Funk [@stapperMindBaseline2025] set out five criteria for a good baseline (capturing key epidemiological dynamics, producing the required output format, using only commonly available data, simplicity, and calibrated uncertainty), and find that no single baseline meets all of them.
:::

# Wrap up

- Review what you've learned in this session with the [learning objectives](../reference/learning_objectives)
- Share your [questions and thoughts](../reference/help)

## References

::: {#refs}
:::
