---
title: "Introduction to the workshop and the instructors"
author: "What is a forecast, and is it any good?"
format:
  revealjs:
    output: slides/introduction-to-the-course-and-the-instructors.html
    footer: "Introduction to the workshop and the instructors"
    chalkboard: true
---

```{r instructors}
instructors <- c("seabbs", "kathsherratt")
```

```{r get_contributors}
library("dplyr")
library("gh")
library("jsonlite")
whoarewe <- lapply(sort(instructors), \(instructor) {
  gh(paste0("GET /users/{username}"), username = instructor)[
    c("login", "avatar_url", "html_url", "name")
  ]
})
```

# Why this workshop?

::: {.incremental}
- forecasts are common in outbreak response and increasingly feed into decisions
- it is easy to make a forecast and much harder to know whether it is any good
- combining forecasts across models often helps, but only if we can measure it
:::

::: {.fragment}
A short, hands-on taster focused on forecasting and whether a forecast is any good, drawn from the fuller [NFIDD course](https://nfidd.github.io/nfidd/).
:::

# Who are we?

```{r whoarewe, results='asis'}
cat(":::: {.columns}\n")
for (instructor in seq_along(whoarewe)) {
  cat(
    "::: {.column width=\"", floor(100 / length(whoarewe)), "%\"}\n",
    "![](", whoarewe[[instructor]]$avatar_url, ")\n",
    "[", whoarewe[[instructor]]$name, "](",
    whoarewe[[instructor]]$html_url, ")\n",
    ":::\n", sep = ""
  )
}
cat("::::\n")
```

# Who are you?

Introduce yourself with:

- name
- where you travelled from
- why you're here

# How the workshop works

::: {.incremental}
- each session opens with a short introduction
- you work through the material hands-on, and we stop at several points to discuss observations and learnings together
- we work throughout from forecasts made earlier, so you can focus on interpreting and evaluating them without waiting for models to fit
- *optional* boxes and the "Going further" sections: skip these on a first pass and come back to them later
:::

#

[Return to the session](../introduction-and-course-overview)
